%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% acknowledgements.tex
%% NOVA thesis document file
%%
%% Text with acknowledgements
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\acknowledgements

%The acknowledgements. You are free to write this section at your own will.  However, usually it starts with the institutional acknowledgements (adviser, institution, grants, workmates, \ldots) and then comes the personal acknowledgements (friends, family, \ldots).

I thank my parents, sister, brother and rest of my family, I am grateful for the love that they share and continuous support. Without them, nothing would be possible. I would like to show my gratitude to all of my friends which stand close to me during my doctoral program, and to my Capoeira family which helped me to become independent and to overcome the cultural difference. 

My sincerest gratitude and appreciation goes to my professors, mentors, and advisers, Dr Vasco Amaral and Dr Miguel Goul\~ao. I am truly indebted for Dr Amaral’s and Dr Goul\~ao involvement in my academic, professional, and social lives for last years. Not only they shepherded me in research, teaching, and outreach
activities, but also gave me an outstanding example, how to become a computer science researcher
while keeping a social consciousness.

I would like to show my gratitude to my NOVA-LINCS research colleagues, which supported me continuously during my research. Especially, to our research group leader Dr Ana Moreira, she was continuously supporting us to work together and share the experiences among colleagues. 

To Dr Marjan Mernik, for his more involved role as a PhD committee member. I really appreciate his advice and direction on each of the key stages in my graduate study. 

I would like to thank Prof. Ademar Aguiar for supporting me in a middle phase of my research by providing me with the opportunity to work and learn on Movercado project and helped me in creating new research ideas. 

I am also indebted to the help and guidance I gained from the Dr Bruno Barroca, Dr Eugene Syriani, Dr Jeff Gray, Dr Hans Vangheluwe, Dr Dominique Blouin and Dr Jacome Cunha. I appreciate their guidance and assistance during my studies.

Finally, I would like to thank the financial support provided by the Department of Computer
Science, of Universidade NOVA de Lisboa, to the department director Dr Luis Caires and my graduate program director Dr Nuno Correia. None of these
can be accomplished without department support.



